---
title: "Course Project 1 - Reproducible Research"
author: "Pankaj Bendre"
date: "March 20, 2016"
output: html_document
---

##Assignment Instructions
1.Code for reading in the dataset and/or processing the data
2.Histogram of the total number of steps taken each day
3.Mean and median number of steps taken each day
4.Time series plot of the average number of steps taken
5.The 5-minute interval that, on average, contains the maximum number of steps
6.Code to describe and show a strategy for imputing missing data
7.Histogram of the total number of steps taken each day after missing values are imputed
8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9.All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

## 1)Loading the Data
```{r,echo=TRUE}
setwd("~/Desktop/R  assignment/Reproducible report/COurse project 1")
activity <- read.csv("activity.csv")
```

## 2) Exploring the basics of data
```{r}
#Loading essential libraries
library(tidyverse)
library(lubridate)
names(activity)
dim(activity)
str(activity)
#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
#transforming the date column into date format using lubridate
activity$date<-ymd(activity$date)
length(unique(activity$date))
```
##Step 2
##Histogram of the total number of steps taken each day
```{r, echo = TRUE}
stepsPerDay <- activity %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsPerDay,10)
#histogram for the total steps for each day
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="blue", xlab="Steps", ylim = c(0,30))

```

##Step 3
## Mean and Median steps taken each day
```{r}
meanPreNA <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianPreNA <- round(median(stepsPerDay$sumsteps),digits = 2)

print(paste("The mean is: ", meanPreNA))

print(paste("The median is: ", medianPreNA))
```
##Step 4
##Time series plot of the average number of steps taken
```{r, echo = TRUE}
stepsPerInterval <- activity %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsPerInterval,10)

#time series plot
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="blue", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")

```
##Step 5
##The 5-minute interval that, on average, contains the maximum number of steps
```{r}
print(paste("Interval containing the most steps on average: ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))

print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```

##Step 6
##6.Code to describe and show a strategy for imputing missing data
```{r}
#Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))
#step for imputing
activityNoNA <- activity  
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){
                activityNoNA$steps[i]<- stepsPerInterval$meansteps[activityNoNA$interval[i] == stepsPerInterval$interval]
        }
}

#After
#Display first 10 rows of data
head(activityNoNA,10)
```

##Step7
##Histogram of the total number of steps taken each day after missing values are imputed
```{r}
stepsPerDay1 <- activityNoNA %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay1,10)
#hist
hist(stepsPerDay1$sumsteps, main = "Histogram of Daily Steps", 
     col="blue", xlab="Steps")
#mean and median steps post imputing NA
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)

print(paste("The mean is: ", mean(meanPostNA)))

print(paste("The median is: ", median(medianPostNA)))

```

##Step 8
## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
#Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)

#Make a panel plot containing a time series plot (i.e. type=“l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)


g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))
```

